Normalization by Evaluation for Sized Dependent Types

### 2017-02-06 nbe/icfp17.tex
[ ICFP 17 ] Before removing term derivation witness from equality.

@@

Syntax (de Bruijn terms).
\[
\begin{array}{lllrl@{\qquad}l}
\Sort & \ni & s
    & ::= & \Set[\ell] ~(\ell \in \NN)& \mbox{sort (universe)} \\
% \Ann & \ni & \star & ::= & \erased \mid \nonstrict \mid \noterased & \mbox{annotation (irrelevant, shape-irrelevant, relevant)}\\
\Ann & \ni & \star & ::= & \erased \mid \noterased & \mbox{annotation (irrelevant, relevant)}\\
\Exp & \ni & t,u,T,U
    & ::= & w \mid t\,e & \mbox{expressions} \\
\Whnf & \ni & w, W
    & ::= & n \mid s \mid \Size \mid \epiT U T % \mid \piSizeT T
    % \mid \forallT T
    \mid \lambda t% & \mbox{functions} \\ &&
    \mid \Nat a \mid c & \mbox{weak head normal forms} \\ % \mbox{natural numbers} \\
\Data & \ni & c
   & ::= & \tzero \mid \suc a t & \mbox{constructed data} \\
\Ne & \ni & n
    & ::= & \ind i \mid n \, e & \mbox{neutral expressions} \\
\Elim & \ni & e
    & ::= & t \mid a \mid \ann a \mid \case[\ell] T {t_1} t_2 \mid \fix[\ell] T t & \mbox{eliminations} \\
\SizeExp & \ni & a,b & ::= & \infty \mid \ind i + o ~(o \in \NN) & \mbox{size expressions} \\
\Cxt & \ni & \Gamma, \Delta & ::= & \cempty \mid \erext[\noterased] \Gamma T \mid \erext \Gamma \Size \\
\end{array}
\]
We omit the ``$\noterased$''-markers from $\Gamma$ and $\Pi$ by default.  We write $\forallT T$ for $\erpiT \Size T$.
For concrete function types, we may use a (named dependent) function type notation as syntactic sugar for the corresponding de Bruijn representation.  For instance, $\funT i \Size \Nat i \to \Set[\ell]$ is sugar for $\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.  We abbreviate this type by \fbox{$\FixK\,\ell$}, and let \fbox{$\FixT\,T$} stand for $\allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$. Similar as for $\Pi$, we use named lambda abstraction as sugar for de Bruijn abstraction.  Named abstract takes care of proper lifting of de Bruijn indices.

@@

We write $\sigma$ and $\tau$ for substitutions, which are lists of terms, and $t \sigma$ for the application of substitution $\sigma$ to term $t$.  We write $\slift^n_m$ for the substitution $(v_{n+m-1},\dots,v_{n+1},v_n)$ which applies to a term with $m$ free indices and increases all of them by $n$.  We write $\slift_m$ for the lifting $\slift^1_m$ and $\sid_m$ for the identity substitution $ \slift^0_m$. The substitution $[u]_m = (\sid_m, u)$ replaces free index $\ind 0$ by term $u$ and decrements all other free indices by $1$. We drop subscript $m$ when clear from the context.

Size increment \fbox{$a + k$} for $k \in \NN$ is defined by $\infty + k = \infty$ and $(\ind i + o) + k = \ind i + (o + k)$.  Size comparison \fbox{$a \leq b$} holds if either $b = \infty$ or $a = \ind i + o$ and $b = \ind i + o'$ with $o \leq o'$.

@@

\fbox{$\Gamma(i) = \eann T$} (implies $\Gamma \der T$).
\[
  \ru{}{(\eext \Gamma T)(0) = \eann T \slift}
\qquad
  \ru{\Gamma(i) = \eann T
    }{(\cext \Gamma \_)(i+1) = \eann T \slift}
\]

@@

\fbox{$\Gamma \der t : T$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der T$ [and $\resurrect \Gamma \der t : T$]. Note: no rule for $\Gamma \der \Size : s$.)
\begin{gather*}
  \ru{\der \Gamma
    }{\Gamma \der \Set[\ell] : \Set[\ell']
    }{\ell {<} \ell'}
\qquad
  \ru{\Gamma \der U : s \qquad
      \cext \Gamma U \der T : s
    }{\Gamma \der \piT U T : s}
\qquad
  \ru{\cext \Gamma \Size \der T : s
    }{\Gamma \der \epiT \Size T : s}
\qquad
  \ru{%\der \Gamma \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \Nat a : s}
\\[2ex]
  \rux{\der \Gamma \qquad \Gamma(i) = \erann[\noterased] T
    }{\Gamma \der \ind i : T
    }{T \not= \Size}
\qquad
  \ru{\eext \Gamma U \der t : T
    }{\Gamma \der \lambda t : \epiT U T}
\qquad
  \ru{\Gamma \der t : \piT U T \qquad
      \Gamma \der u : U
    }{\Gamma \der t\,u : T[u]}
% \qquad
%   \ru{\Gamma \der t : T \qquad
%       \Gamma \mid T \der e : T'
%     }{\Gamma \der t\,e : T'}
\qquad
  \ru{\Gamma \der t : \piT \Size T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der t\,a : T[a]}
\\[2ex] % \qquad
  \ru{\resurrect\Gamma \der b : \Size
    }{\Gamma \der \tzero : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der a : \Size \qquad
      \Gamma \der t : \Nat b
    }{\Gamma \der \suc a t : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der t : \forallT T \qquad
      \Gamma \der a : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der t \ann a : T[b]}
\\[2ex]
  \ru{\Gamma \der u : \Nat (b + 1) \qquad
      \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
      \Gamma \der t_z : T \, \tzero \qquad
      \Gamma \der a : \Size \qquad
      \Gamma \der t_s : (n : \Nat b) \to T \, (\suc a n)
    }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
\\[2ex]
  \ru{\Gamma \der u : \Nat b \qquad
      \Gamma \der T : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t : \FixT\, T
     }{\Gamma \der u\,\fix[\ell] T t : T \,b \, u}
\end{gather*}
Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
% \fbox{$\Gamma \mid U \der e : T$}
% \begin{gather*}
%   \ru{\cext \Gamma U \der T \qquad \Gamma \der u : U
%     }{\Gamma \mid \piT U T \der u : T[u]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \piT \Size T \der a : T[a]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \forallT T \der \ann a : T[a]}
% \\
%   \ru{
%     }{\Gamma \mid \Nat a \der \fix T t : T \ann a n}
% \end{gather*}

@@

@@

@@

@@

@@

>>>>

\usepackage[cal=boondoxo]{mathalfa}

@@

Syntax (de Bruijn terms).
\[
\begin{array}{lllrl@{\qquad}l}
\Sort & \ni & s
    & ::= & \Set[\ell] ~(\ell \in \NN)& \mbox{sort (universe)} \\
% \Ann & \ni & \star & ::= & \erased \mid \nonstrict \mid \noterased & \mbox{annotation (irrelevant, shape-irrelevant, relevant)}\\
\Ann & \ni & \star & ::= & \erased \mid \noterased & \mbox{annotation (irrelevant, relevant)}\\
\Exp & \ni & t,u,T,U
    & ::= & w \mid t\,e & \mbox{expressions} \\
\Whnf & \ni & w, W
    & ::= & n \mid s \mid \Size \mid \epiT U T % \mid \piSizeT T
    % \mid \forallT T
    \mid \lambda t% & \mbox{functions} \\ &&
    \mid \Nat a \mid c & \mbox{weak head normal forms} \\ % \mbox{natural numbers} \\
\Data & \ni & c
   & ::= & \tzero \mid \suc a t & \mbox{constructed data} \\
\Ne & \ni & n
    & ::= & \ind i \mid n \, e & \mbox{neutral expressions} \\
\Elim & \ni & e
    & ::= & t \mid a \mid \ann a \mid \case[\ell] T {t_1} t_2 \mid \fix[\ell] T t & \mbox{eliminations} \\
\SizeExp & \ni & a,b & ::= & \infty \mid k \mid \ind i + k ~(k \in \NN) & \mbox{size expressions} \\
\Cxt & \ni & \Gamma, \Delta & ::= & \cempty \mid \erext[\noterased] \Gamma T \mid \erext \Gamma \Size \\
\end{array}
\]
We omit the ``$\noterased$''-markers from $\Gamma$ and $\Pi$ by default.  We write $\forallT T$ for $\erpiT \Size T$.
For concrete function types, we may use a (named dependent) function type notation as syntactic sugar for the corresponding de Bruijn representation.  For instance, $\funT i \Size \Nat i \to \Set[\ell]$ is sugar for $\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.  We abbreviate this type by \fbox{$\FixK\,\ell$}, and let \fbox{$\FixT\,T$} stand for $\allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$. Similar as for $\Pi$, we use named lambda abstraction as sugar for de Bruijn abstraction.  Named abstract takes care of proper lifting of de Bruijn indices, for instance, $\lambda x.\, t x = \lambda. (t \slift)\,\ind 0$ if $t$ is outside the scope of $x$.

@@

We write $\sigma$ and $\tau$ for substitutions, which are lists of terms, and $t \sigma$ for the application of substitution $\sigma$ to term $t$.  We write $\slift^k_m$ for the substitution $(v_{k+m-1},\dots,v_{k+1},v_k)$ which applies to a term with $m$ free indices and increases all of them by $k$.  We write $\slift_m$ for the lifting $\slift^1_m$ and $\sid_m$ for the identity substitution $ \slift^0_m$. The substitution $[u]_m = (\sid_m, u)$ replaces free index $\ind 0$ by term $u$ and decrements all other free indices by $1$. We drop subscript $m$ when clear from the context.

Size increment \fbox{$a + k$} for $k \in \NN$ extends addition by $\infty + k = \infty$ and $(\ind i + k') + k = \ind i + (k' + k)$.  Size comparison \fbox{$a \leq b$} holds if either $b = \infty$ or  $k \leq k'$ where either $a = k$ and $b = k'$ or $a \in \{ k, \ind i + k \}$ and $b = \ind i + k'$.

@@

\fbox{$\Gamma(i) = \eann T$} (implies $\resurrect\Gamma \der T$).
\[
  \ru{}{(\eext \Gamma T)(0) = \eann T \slift}
\qquad
  \ru{\Gamma(i) = \eann T
    }{(\cext \Gamma \_)(i+1) = \eann T \slift}
\]

@@

\fbox{$\Gamma \der t : T$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der T$ [and $\resurrect \Gamma \der t : T$]. Note: no rule for $\Gamma \der \Size : s$.)
\begin{gather*}
  \ru{\der \Gamma
    }{\Gamma \der \Set[\ell] : \Set[\ell']
    }{\ell {<} \ell'}
\qquad
  \ru{\Gamma \der U : s \qquad
      \cext \Gamma U \der T : s
    }{\Gamma \der \piT U T : s}
\qquad
  \ru{\cext \Gamma \Size \der T : s
    }{\Gamma \der \epiT \Size T : s}
\qquad
  \ru{%\der \Gamma \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \Nat a : s}
\\[2ex]
  \rux{\der \Gamma \qquad \Gamma(i) = \erann[\noterased] T
    }{\Gamma \der \ind i : T
    }{T \not= \Size}
\qquad
  \ru{\eext \Gamma U \der t : T
    }{\Gamma \der \lambda t : \epiT U T}
\qquad
  \ru{\Gamma \der t : \piT U T \qquad
      \Gamma \der u : U
    }{\Gamma \der t\,u : T[u]}
% \qquad
%   \ru{\Gamma \der t : T \qquad
%       \Gamma \mid T \der e : T'
%     }{\Gamma \der t\,e : T'}
\qquad
  \ru{\Gamma \der t : \piT \Size T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der t\,a : T[a]}
\\[2ex] % \qquad
  \ru{\resurrect\Gamma \der b : \Size
    }{\Gamma \der \tzero : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der a : \Size \qquad
      \Gamma \der t : \Nat b
    }{\Gamma \der \suc a t : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der t : \forallT T \qquad
      \Gamma \der a : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der t \ann a : T[b]}
\\[2ex]
  \ru{\Gamma \der u : \Nat (b + 1) \qquad
      \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
      \Gamma \der t_z : T \, \tzero \qquad
      \Gamma \der a : \Size \qquad
      \Gamma \der t_s : (t : \Nat b) \to T \, (\suc a t)
    }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
\\[2ex]
  \ru{\Gamma \der u : \Nat b \qquad
      \Gamma \der T : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t : \FixT\, T
     }{\Gamma \der u\,\fix[\ell] T t : T \,b \, u}
\end{gather*}
Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
% \fbox{$\Gamma \mid U \der e : T$}
% \begin{gather*}
%   \ru{\cext \Gamma U \der T \qquad \Gamma \der u : U
%     }{\Gamma \mid \piT U T \der u : T[u]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \piT \Size T \der a : T[a]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \forallT T \der \ann a : T[a]}
% \\
%   \ru{
%     }{\Gamma \mid \Nat a \der \fix T t : T \ann a n}
% \end{gather*}

@@

      \Gamma \der t_s : (t : \Nat b) \to T\,(\suc a t)

@@

\begin{lemma}[Inversion of typing]
\label{lem:invty} \bla
\begin{enumerate}
\item TODO
\end{enumerate}
\end{lemma}

@@

\subsection{Operational semantics}

@@

We write \fbox{$t \approx t'$} iff $t$ and $t'$ are syntactically equal if we ignore sizes $\ann a$ in irrelevant positions.  For example, $\suc a \tzero \approx \suc 0 \tzero$, but $\Nat a \not\approx \Nat 0$.

\subsection{Logical predicates}

We write $\forall \rho : \Gamma \leq \Delta.~ P(\Gamma,\rho)$ short for
$\forall \Gamma, \rho.~ \Gamma \der \rho : \Delta \implies P(\Gamma,\rho)$.
We apply implicit binding (like $\Gamma$ here) in other, analogous cases as well.

We write \fbox{$\Gamma \der t :=: t' : T$} for the conjunction of $\Gamma \der t,t' : T$ and $\Gamma \der t = t' : T$.  We write \fbox{$\Gamma \der t \evalsto w : T$} if additionally $t \evalsto w$.

We define a family of PERs \fbox{$\sneq \Delta \_ \_ T$} that characterize the neutrals of type $T$.
\[
  \sneq \Delta t {t'} T \miff \Delta \der t \evalsto n : T \mand \Delta \der t' \evalsto n' : T
\]

Assuming $\Delta \der b : \Size$, we define inductively a family \fbox{$\snateq \Delta \_ \_ b$} of relations that characterize the semantic natural numbers.
\begin{gather*}
  \ru{\sneq \Delta t {t'} {\Nat b}
    }{\snateq \Delta t {t'} b
    }
\\[2ex]
  \rux{\Delta \der t \evalsto \tzero : \Nat b \qquad
      \Delta \der t' \evalsto \tzero : \Nat b
    }{\snateq \Delta t {t'} b
    }{b > 0}
\\[2ex]
  \rux{\Delta \der t \evalsto \suc a u : \Nat b \qquad
      \Delta \der t' \evalsto \suc{a'}{u'} : \Nat b \qquad
      \snateq \Delta u {u'} {b'}
    }{\snateq \Delta t {t'} b
    }{b = b' + 1}
\end{gather*}
Note that in the last rule $b$ and $b'$ could be both $\infty$.
We write \fbox{$\snat \Delta t b$} iff $\snateq \Delta t t b$.

By recursion on level $\ell \in \NN$ we define inductively what it means to be a semantic type, \fbox{$\sty \Delta T \ell$}, and simultaneously
\begin{enumerate}
\item what it means for semantic types to be equal, \fbox{$\steqd \Delta T {T'} \ell {\T}$},
\item what it means for a term to have a semantic type, \fbox{$\tcal :: \sttmd \Delta t T {\T}$}, and
\item what it means for two such terms to be equal, \fbox{$\sttmeqd \Delta t {t'} T {\T} {\tcal}$},
\end{enumerate}
all three by recursion on a derivation $\T :: \sty \Delta T \ell$.
For each $\ell$, this is an instance of an inductive-recursive definition \cite{dybjer:jsl00}.

We first list the rules for \fbox{$\sty \Delta T \ell$}:
\begin{gather*}
  \rux{\Delta \der T \evalsto \Set[\ell'] : \Set[\ell]
     }{\sty \Delta T \ell
     }{\ell' {<} \ell}
\qquad
  \ru{\Delta \der T \evalsto N : \Set[\ell]
    }{\sty \Delta T \ell}
\qquad
  \ru{\Delta \der T \evalsto \Nat a : \Set[\ell]
    }{\sty \Delta T \ell}
\\[2ex]
  \rul{\Delta \der T' \evalsto \epiT \Size T : \Set[\ell] \qquad
       \eext \Delta \Size \der T : \Set[\ell] \qquad
      \forall \rho : \Gamma \leq \Delta,~\Gamma \der a : \Size.~
        \sty \Gamma {T(\rho,a)} \ell \\
    }{\sty \Delta {T'} \ell}
\\[2ex]
  \rul{\Delta \der T' \evalsto \piT U T : \Set[\ell] \qquad \cext \Delta U \der T : \Set[\ell] \\ \relax
      \U :: \forall \rho : \Gamma \leq \Delta.~ \sty{\Gamma}{U\rho} {\ell} \\ \relax
      \T :: \forall \rho : \Gamma \leq \Delta,~\sttmd \Gamma u {U\rho} {\U(\rho)}.~
        \sty \Gamma {T(\rho,u)} \ell \\
      \forall \rho : \Gamma \leq \Delta,~ \uu :: \sttmd \Gamma u {U\rho} {\U(\rho)}.~
        \sttmeqd \Gamma u {u'} {U\rho} {\U(\rho)} {\uu} \implies
        \steqd \Gamma {T(\rho,u)} {T(\rho,u')} \ell {\T(\rho,\uu)}
    }{\sty \Delta {T'} \ell}
\end{gather*}
Then, we give the cases for \fbox{$\steqd \Delta T {T'} \ell {\T}$}.
\begin{caselist}

\nextcase Universes:
\[
  \rux{\Delta \der T \evalsto \Set[\ell'] : \Set[\ell]
     }{\sty \Delta T \ell
     }{\ell' {<} \ell}
\]
Then $\steqd \Delta T {T'} \ell {\T}$ iff $\Delta \der T' \evalsto \Set[\ell'] : \Set[\ell]$.

\nextcase Neutrals:
\[
  \ru{\Delta \der T \evalsto N : \Set[\ell]
    }{\sty \Delta T \ell}
\]
Then $\steqd \Delta T {T'} \ell {\T}$ iff $\Delta \der T' \evalsto N' : \Set[\ell]$ and
$\Delta \der N = N' : \Set[\ell]$.

\nextcase Natural number type:
\[
  \ru{\Delta \der T \evalsto \Nat a : \Set[\ell]
    }{\sty \Delta T \ell}
\]
Then $\steqd \Delta T {T'} \ell {\T}$ iff $\Delta \der T' \evalsto \Nat a : \Set[\ell]$.

\nextcase Size quantification:
\[
  \rul{\Delta \der T \evalsto \epiT \Size U : \Set[\ell] \qquad
       \eext \Delta \Size \der U : \Set[\ell] \qquad
      \U :: \forall \rho : \Gamma \leq \Delta,~\Gamma \der a : \Size.~
        \sty \Gamma {U(\rho,a)} \ell \\
    }{\sty \Delta {T} \ell}
\]
Then $\steqd \Delta T {T'} \ell {\T}$ iff $\Delta \der T' \evalsto \epiT \Size {U'} : \Set[\ell]$
and $\eext \Delta \Size \der U' : \Set[\ell]$ and forall $\rho : \Gamma \leq \Delta$ and $\Gamma \der a : \Size$ we have $\steqd \Gamma {U(\rho,a)} {U'(\rho,a)} \ell {\U(\rho,a)}$.

\nextcase Function spaces:
\[
  \rul{\Delta \der T_1' \evalsto \piT {U_1} {T_1} : \Set[\ell] \qquad \cext \Delta U_1 \der T_1 : \Set[\ell] \\ \relax
      \U :: \forall \rho : \Gamma \leq \Delta.~ \sty{\Gamma}{U_1\rho} {\ell} \\ \relax
      \T :: \forall \rho : \Gamma \leq \Delta,~\sttmd \Gamma u {U_1\rho} {\U(\rho)}.~
        \sty \Gamma {T_1(\rho,u)} \ell \\
      \forall \rho : \Gamma \leq \Delta,~ \uu :: \sttmd \Gamma u {U_1\rho} {\U(\rho)}.~
        \sttmeqd \Gamma u {u'} {U_1\rho} {\U(\rho)} {\uu} \implies
        \steqd \Gamma {T_1(\rho,u)} {T_1(\rho,u')} \ell {\T(\rho,\uu)}
    }{\sty \Delta {T_1'} \ell}
\]
Then $\steqd \Delta {T_1'} {T_2'} \ell {\T}$ iff $\Delta \der T_2' \evalsto \piT{U_2}{T_2} : \Set[\ell]$ and
$\cext \Delta{U_2} \der T_2 : \Set[\ell]$ and
forall $\rho : \Gamma \leq \Delta$ we have $\steqd \Gamma {U_1\rho} {U_2\rho} \ell {\U(\rho)}$
and forall $\uu :: \sttmd \Gamma u {U_1\rho} {\U(\rho)}$ finally
$\steqd \Gamma {T_1(\rho,u)} {T_2(\rho,u)} \ell {\T(\rho,\uu)}$.

\end{caselist}

Next, we define semantic membership \fbox{$\sttmd \Delta t T {\T}$}.
\begin{caselist}

\nextcase
\[
  \rux{\Delta \der T \evalsto \Set[\ell'] : \Set[\ell]
     }{\sty \Delta T \ell
     }{\ell' {<} \ell}
\]
Then $\sttmd \Delta t T {\T}$ iff $\sty \Delta t {\ell'}$.

\nextcase
\[
  \ru{\Delta \der T \evalsto N : \Set[\ell]
    }{\sty \Delta T \ell}
\]
Then $\sttmd \Delta t T {\T}$ iff $\Delta \der t \evalsto n : N$.

\nextcase
\[
  \ru{\Delta \der T \evalsto \Nat a : \Set[\ell]
    }{\sty \Delta T \ell}
\]
Then $\sttmd \Delta t T {\T}$ iff $\snat \Delta t a$.

\nextcase %  Let $\T' :: \sty \Delta {T'} \ell$ be derived by
\[
  \T' ::
  \rul{\Delta \der T' \evalsto \piT \Size T : \Set[\ell] \qquad
       \cext \Delta \Size \der T : \Set[\ell] \qquad
      \T :: \forall \rho : \Gamma \leq \Delta,~\Gamma \der a : \Size.~
        \sty \Gamma {T(\rho,a)} \ell \\
    }{\sty \Delta {T'} \ell}
\]
Then $\sttmd \Delta t {T'} {\T'}$ iff $\Delta \der t : T'$ and
forall $\rho : \Gamma \leq \Delta$ and $\acal :: \Gamma \der a : \Size$
we have $\sttmd \Gamma {t\,a} {T(\rho,a)} {\T(\rho,\acal)}$.

\nextcase % Let $\T' :: \sty \Delta {T'} \ell$ be derived by
\[
  \T' ::
  \rul{\Delta \der T' \evalsto \erpiT \Size T : \Set[\ell] \qquad
       \erext \Delta \Size \der T : \Set[\ell] \qquad
      \T :: \forall \rho : \Gamma \leq \Delta,~\Gamma \der b : \Size.~
        \sty \Gamma {T(\rho,b)} \ell \\
    }{\sty \Delta {T'} \ell}
\]
Then $\sttmd \Delta t {T'} {\T'}$ iff $\Delta \der t : T'$ and
forall $\rho : \Gamma \leq \Delta$ and $\bb :: \Gamma \der b : \Size$ and  $\resurrect\Gamma \der a : \Size$,
\begin{enumerate}
\item $\tcal :: \sttmd \Gamma {t \ann a} {T(\rho,b)} {\T(\rho,\bb)}$, and
\item $\sttmeqd \Gamma {t \ann a} {t \ann {a'}} {T(\rho,b)} {\T(\rho,\bb)} {\tcal}$ for all $\resurrect\Gamma \der a' : \Size$.
\end{enumerate}

\nextcase
\[
  \T' ::
  \rul{\Delta \der T' \evalsto \piT U T : \Set[\ell] \qquad \cext \Delta U \der T : \Set[\ell] \\ \relax
      \U :: \forall \rho : \Gamma \leq \Delta.~ \sty{\Gamma}{U\rho} {\ell} \\ \relax
      \T :: \forall \rho : \Gamma \leq \Delta,~\sttmd \Gamma u {U\rho} {\U(\rho)}.~
        \sty \Gamma {T(\rho,u)} \ell \\
      \forall \rho : \Gamma \leq \Delta,~ \uu :: \sttmd \Gamma u {U\rho} {\U(\rho)}.~
        \sttmeqd \Gamma u {u'} {U\rho} {\U(\rho)} {\uu} \implies
        \steqd \Gamma {T(\rho,u)} {T(\rho,u')} \ell {\T(\rho,\uu)}
    }{\sty \Delta {T'} \ell}
\]
Then $\sttmd \Delta t {T'} {\T'}$ iff $\Delta \der t : T'$ and
forall $\rho : \Gamma \leq \Delta$ and $\uu :: \sttmd \Gamma u {U\rho} {\U(\rho)}$,
\begin{enumerate}
\item $\tcal :: \sttmd \Gamma {t\rho\,u} {T(\rho,u)} {\T(\rho,\uu)}$, and
\item $\sttmeqd \Gamma {t\rho\,u} {t\rho\,u'} {T(\rho,u)} {\T(\rho,\uu)} {\tcal}$ for all
  $\sttmeqd \Gamma u {u'} {U\rho} {\U(\rho)} {\uu}$.
\end{enumerate}

\end{caselist}

Finally, we define \fbox{$\sttmeqd \Delta t {t'} T {\T} {\tcal}$}
where $\tcal :: \sttmd \Delta t T {\T}$.
\begin{caselist}

\nextcase
\[
  \T ::
  \rux{\Delta \der T \evalsto \Set[\ell'] : \Set[\ell]
     }{\sty \Delta T \ell
     }{\ell' {<} \ell}
\]
Then $\sttmeqd \Delta t {t'} T {\T} {\tcal}$ iff
% $\T' : \sty \Delta {t} \ell$ and
$\steqd \Delta t {t'} {\ell'} {\tcal}$.

\nextcase
\[
  \ru{\Delta \der T \evalsto N : \Set[\ell]
    }{\sty \Delta T \ell}
\]
Then $\sttmeqd \Delta t {t'}T {\T}$ iff $\Delta \der t \evalsto n : N$.

\nextcase
\[
  \ru{\Delta \der T \evalsto \Nat a : \Set[\ell]
    }{\sty \Delta T \ell}
\]
Then $\sttmd \Delta t T {\T}$ iff $\snat \Delta t a$.

\nextcase %  Let $\T' :: \sty \Delta {T'} \ell$ be derived by
\[
  \T' ::
  \rul{\Delta \der T' \evalsto \piT \Size T : \Set[\ell] \qquad
       \cext \Delta \Size \der T : \Set[\ell] \qquad
      \T :: \forall \rho : \Gamma \leq \Delta,~\Gamma \der a : \Size.~
        \sty \Gamma {T(\rho,a)} \ell \\
    }{\sty \Delta {T'} \ell}
\]
Then $\sttmd \Delta t {T'} {\T'}$ iff $\Delta \der t : T'$ and
forall $\rho : \Gamma \leq \Delta$ and $\acal :: \Gamma \der a : \Size$
we have $\sttmd \Gamma {t\,a} {T(\rho,a)} {\T(\rho,\acal)}$.

\nextcase % Let $\T' :: \sty \Delta {T'} \ell$ be derived by
\[
  \T' ::
  \rul{\Delta \der T' \evalsto \erpiT \Size T : \Set[\ell] \qquad
       \erext \Delta \Size \der T : \Set[\ell] \qquad
      \T :: \forall \rho : \Gamma \leq \Delta,~\Gamma \der b : \Size.~
        \sty \Gamma {T(\rho,b)} \ell \\
    }{\sty \Delta {T'} \ell}
\]
Then $\sttmd \Delta t {T'} {\T'}$ iff $\Delta \der t : T'$ and
forall $\rho : \Gamma \leq \Delta$ and $\bb :: \Gamma \der b : \Size$ and  $\resurrect\Gamma \der a : \Size$,
\begin{enumerate}
\item $\tcal :: \sttmd \Gamma {t \ann a} {T(\rho,b)} {\T(\rho,\bb)}$, and
\item $\sttmeqd \Gamma {t \ann a} {t \ann {a'}} {T(\rho,b)} {\T(\rho,\bb)} {\tcal}$ for all $\resurrect\Gamma \der a' : \Size$.
\end{enumerate}

\nextcase
\[
  \T' ::
  \rul{\Delta \der T' \evalsto \piT U T : \Set[\ell] \qquad \cext \Delta U \der T : \Set[\ell] \\ \relax
      \U :: \forall \rho : \Gamma \leq \Delta.~ \sty{\Gamma}{U\rho} {\ell} \\ \relax
      \T :: \forall \rho : \Gamma \leq \Delta,~\sttmd \Gamma u {U\rho} {\U(\rho)}.~
        \sty \Gamma {T(\rho,u)} \ell \\
      \forall \rho : \Gamma \leq \Delta,~ \uu :: \sttmd \Gamma u {U\rho} {\U(\rho)}.~
        \sttmeqd \Gamma u {u'} {U\rho} {\U(\rho)} {\uu} \implies
        \steqd \Gamma {T(\rho,u)} {T(\rho,u')} \ell {\T(\rho,\uu)}
    }{\sty \Delta {T'} \ell}
\]
Then $\sttmd \Delta t {T'} {\T'}$ iff $\Delta \der t : T'$ and
forall $\rho : \Gamma \leq \Delta$ and $\uu :: \sttmd \Gamma u {U\rho} {\U(\rho)}$,
\begin{enumerate}
\item $\tcal :: \sttmd \Gamma {t\rho\,u} {T(\rho,u)} {\T(\rho,\uu)}$, and
\item $\sttmeqd \Gamma {t\rho\,u} {t\rho\,u'} {T(\rho,u)} {\T(\rho,\uu)} {\tcal}$ for all
  $\sttmeqd \Gamma u {u'} {U\rho} {\U(\rho)} {\uu}$.
\end{enumerate}

\end{caselist}

@@

\section{Metatheory}
\label{sec:meta}

\begin{lemma}[Typing of constructed data]
\label{lem:invdata} \bla
If\/ $\Gamma \der c : \Nat b$ then $\resurrect\Gamma \der b : \Size$ and $b > 0$.
\end{lemma}
